Supplementary information: Variational theory of the ground state of quantum electrodynamics

### 2018-10-16 supp.tex

\title{Supplementary information for: Variational theory of the ground state of quantum electrodynamics}

@@

%\begin{align}
%H_{el} = &\int d^3x ~\psi^{\dagger}(x)\left(-\frac{\hbar^2\nabla^2}{2m} + v_{ext} \right)\psi(x) \nonumber \\ &+ \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(x)\psi^{\dagger}(x')V(x-x')\psi(x')\psi(x),
%\end{align}
\begin{equation}
H_{el} = \int d^3x ~\psi^{\dagger}(x)\left(-\frac{\hbar^2\nabla^2}{2m} + v_{ext} \right)\psi(x) + \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(x)\psi^{\dagger}(x')V(x-x')\psi(x')\psi(x),
\end{equation}
where $v_{ext}$ is the one-body potential, $V(x-x')$ is the two-body interaction kernel, and $\psi$ is the second-quantized electron field.

@@

The interaction Hamiltonian takes the form:
%\begin{align}
%H_{int} = &\frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(x)(\mathbf{A}\cdot\nabla +  \nabla \cdot \mathbf{A})\psi(x) + \nonumber \\ &\frac{e^2}{2m}\int d^3x ~\psi^{\dagger}\psi\mathbf{A}^2(x).
%\end{align}
\begin{equation}
H_{int} = \frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(x)(\mathbf{A}\cdot\nabla +  \nabla \cdot \mathbf{A})\psi(x) + \frac{e^2}{2m}\int d^3x ~\psi^{\dagger}\psi\mathbf{A}^2(x).
\end{equation}

@@

The last thing to note before taking the expectation values and minimization is that we should focus on the submanifold of variations that keep the complete set of electron and photon functions normalized. This shall be enforced by constructing the Lagrange function:
\begin{widetext}
\begin{equation}
\mathcal{L}[\{ \psi_n,\psi_n^* \},\{ \mathbf{F}_i,\mathbf{F}_i^*,\omega_i \}] = \langle \Omega |H|\Omega\rangle - \sum_n E_n\left(\int d^3x ~\psi_n^*\psi_n - 1 \right) - \sum_n \frac{\hbar\lambda_i}{2}\left(\int d^3x ~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_i - 1 \right).
\end{equation}
\end{widetext}

@@

The minimization with respect to the electron orbital $\psi_i^*$ yields the equation:
% \begin{widetext}
% \begin{equation}
% \left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) +  \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\left(\psi^*_j(x')\psi_j(x')\psi_i(x) - \psi_j(x')\psi_j(x)\psi_i(x')  \right) + \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x)  = E_i\psi_i(x).
% \end{equation}
% \end{widetext}
\begin{align}
\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) &+  \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\left(\psi^*_j(x')\psi_j(x')\psi_i(x) - \psi_j(x')\psi_j(x)\psi_i(x')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x)  = E_i\psi_i(x).
\end{align}
This differs from the usual Hartree-Fock equation of electronic theory by the last term on the left-hand-side, representing a polarization induced by the $\mathbf{A}^2$ term.

@@

In the derivation of Equation (11), it is notable that the term linear in the vector potential vanishes. In many quantum electrodynamical problems, this linear term is important. To first order, it leads to spontaneous emission of a photon. To second order, it leads to van der Waals / Casimir-Polder forces on emitters, as well as effective interactions betweeen distinct emitters. Thus, we seek to capture the effect of this term. Physically, this term will mix the factorizable ground state of Equation (7) with terms that have virtual excitations of the matter, as well as virtual excitations of the electromagnetic field. The resulting state is now non-factorizable and we thus conclude that the term in the Hamiltonian linear in the vector potential leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of the quantum electrodynamical ground state.

We capture the effect of correlations perturbatively. In other words, we consider the second-order correction $\delta E$ to the ground state energy arising from the term in the Hamiltonian linear in the vector potential. That correction is given by
\begin{equation}
\delta E = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{q=1}^{N_p}\sum_{m=N_{\sigma}+1}^{\infty}\sum\limits_{1}^{N_{\sigma}} \frac{\Big| \int d^3x~\mathbf{F}_q^*\cdot\mathbf{j}_{nm}\Big|^2}{\omega_q(\omega_{mn} -\lambda_q)},
\end{equation}
where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$,$\omega_{mn} = \omega_m - \omega_n$, $N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons (divided by 2 if spin is retained), and $N_p$ is the number of photon modes retained.  In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Equations (11) and (14), and then all the information necessary to calculate the correlation energy is present. In what follows, we add this energy correction $\delta E$ to the expectation value of the energy in the ground state, with the orbitals and eigenvalues as variational parameters. As a result, the orbitals and eigenvalues obtained will be different from Equation (11) and (14), and this difference will be small in the limit of weak correlations. Strictly speaking, this approach is only justified for weak correlations, but can be applied to systems with strong-correlations as is often done with self-consistent methods.

@@

The derivative of $\delta E$ with respect to an occupied electron orbital $k$ is given by
\begin{equation}
\frac{\partial\delta E}{\partial \psi_k^*} = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{N_p} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{nk}(\mathbf{y})}{\omega_q(\omega_{kn}-\lambda_q)}\int d^3x~\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)
\end{equation}
A similar equation arises for the derivative with respect to an unoccupied electron orbital, except that the summation should now be over unoccupied electron orbitals.

>>>>

\title{Supplementary information: Variational theory of the ground state of quantum electrodynamics}

@@

%\begin{align}
%H_{el} = &\int d^3x ~\psi^{\dagger}(x)\left(-\frac{\hbar^2\nabla^2}{2m} + v_{ext} \right)\psi(x) \nonumber \\ &+ \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(x)\psi^{\dagger}(x')V(x-x')\psi(x')\psi(x),
%\end{align}
\begin{equation}
H_{el} = \int d^3x ~\psi^{\dagger}(x)\left(-\frac{\hbar^2\nabla^2}{2m} + v_{ext}\Jadd{(x)} \right)\psi(x) + \frac{1}{2}\int d^3x d^3x'~ \psi^{\dagger}(x)\psi^{\dagger}(x')V(x-x')\psi(x')\psi(x),
\end{equation}
where $v_{ext}$ is the one-body potential, $V(x-x')$ is the two-body interaction kernel, and $\psi$ is the second-quantized electron field. \Jadd{(shouldn't x also be bold when it is an argument, since it a vector)}

@@

The interaction Hamiltonian takes the form:
%\begin{align}
%H_{int} = &\frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(x)(\mathbf{A}\cdot\nabla +  \nabla \cdot \mathbf{A})\psi(x) + \nonumber \\ &\frac{e^2}{2m}\int d^3x ~\psi^{\dagger}\psi\mathbf{A}^2(x).
%\end{align}
\begin{equation}
H_{int} = \frac{-i\hbar e}{2m}\int d^3x ~\psi^{\dagger}(x)(\mathbf{A}\Jadd{(x)}\cdot\nabla +  \nabla \cdot \mathbf{A}\Jadd{(x)})\psi(x) + \frac{e^2}{2m}\int d^3x ~\psi^{\dagger}\Jadd{(x)}\psi\Jadd{(x)}\mathbf{A}^2(x).
\end{equation}

@@

The last thing to note before taking the expectation values and minimization is that we should focus on the submanifold of variations that keep the complete set of electron and photon functions normalized. This shall be enforced by constructing the Lagrange function:
\begin{widetext}
\begin{equation}
\mathcal{L}[\{ \psi_n,\psi_n^* \},\{ \mathbf{F}_i,\mathbf{F}_i^*,\omega_i \}] = \langle \Omega |H|\Omega\rangle - \sum_n E_n\left(\int d^3x ~\psi_n^*\psi_n - 1 \right) - \sum_n \frac{\hbar\lambda_i}{2}\left(\int d^3x ~\epsilon\mathbf{F}_i^*\cdot\mathbf{F}_i - 1 \right).
\end{equation}
\end{widetext}
\Jadd{(say where the $\lambda$ is coming from)}

@@

The minimization with respect to the electron orbital $\psi_i^*$ \Jadd{($\delta\mathcal{L}/\delta\psi_i^*=0$) }yields the equation:
% \begin{widetext}
% \begin{equation}
% \left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) +  \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\left(\psi^*_j(x')\psi_j(x')\psi_i(x) - \psi_j(x')\psi_j(x)\psi_i(x')  \right) + \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n|^2\psi_i(x)  = E_i\psi_i(x).
% \end{equation}
% \end{widetext}
\begin{align}
\left(\frac{\mathbf{p}^2}{2m}+v_{ext}(x) \right)\psi_i(x) &+  \sum\limits_{j=1}^N \int d^3x' ~ V(x-x')\left(\psi^*_j(x')\psi_j(x')\psi_i(x) - \psi_j(x')\psi_j(x)\psi_i(x')  \right) \nonumber \\ &+ \frac{\hbar e^2}{4m\epsilon_0}\sum_n \frac{1}{\omega_n}|\mathbf{F}_n\Jadd{(x)}|^2\psi_i(x)  = E_i\psi_i(x).
\end{align}
This differs from the usual Hartree-Fock equation of electronic theory by the last term on the left-hand-side, representing a polarization induced by the $\mathbf{A}^2$ term.

@@

In the derivation of Equation (11), it is notable that the term linear in the vector potential vanishes. In many quantum electrodynamical problems, this linear term is important. To first order, it leads to spontaneous emission of a \Jadd{single} photon. To second order, it leads to \Jadd{two-photon processes, such as} van der Waals / Casimir-Polder forces on emitters, as well as effective interactions betweeen distinct emitters. Thus, we seek to capture the effect of this term. Physically, this term will mix the factorizable ground state of Equation (7) with terms that have virtual excitations of the matter, as well as virtual excitations of the electromagnetic field. The resulting state is now non-factorizable and we thus conclude that the term in the Hamiltonian linear in the vector potential leads to \textit{correlations} in the system, and contributes wholly at lowest order to the correlation energy of the quantum electrodynamical ground state.

We capture the effect of correlations perturbatively. In other words, we consider the second-order correction $\delta E$ to the ground state energy arising from the term in the Hamiltonian linear in the vector potential. That correction is given by
\begin{equation}
\delta E = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{q=1}^{N_p}\sum_{m=N_{\sigma}+1}^{\infty}\sum\limits_{1}^{N_{\sigma}} \frac{\Big| \int d^3x~\mathbf{F}_q^*\cdot\mathbf{j}_{nm}\Big|^2}{\omega_q(\omega_{mn} -\lambda_q)},
\end{equation}
where $\mathbf{j}_{nm} = \psi^*_n\nabla\psi_m - (\nabla\psi^*_n)\psi_m$, $\omega_{mn} = \omega_m - \omega_n$, $N_{\sigma}$ is the number of occupied orbitals, equal to the number of electrons (divided by 2 if spin is retained), and $N_p$ is the number of photon modes retained.  In a method without self-consistency, the electron and photon orbitals and eigenvalues are those obtained from Equations (11) and (14), and then all the information necessary to calculate the correlation energy is present. In what follows, we add this energy correction $\delta E$ to the expectation value of the energy in the ground state, with the orbitals and eigenvalues as variational parameters. As a result, the orbitals and eigenvalues obtained will be different from Equation (11) and (14), and this difference will be small in the limit of weak correlations. Strictly speaking, this approach is only justified for weak correlations, but can be applied to systems with strong-correlations as is often done with self-consistent methods.

@@

The derivative of $\delta E$ with respect to an occupied electron orbital $k$ is given by \Jadd{(here you have the y, and x bold, best if also do so above)}
\begin{equation}
\frac{\partial\delta E}{\partial \psi_k^*} = \frac{e^2\hbar^2}{8m^2\epsilon_0}\sum\limits_{n=N_{\sigma}+1}^{\infty}\sum\limits_{q=1}^{N_p} \frac{\int d^3y~\mathbf{F}^*_q(\mathbf{y})\cdot\mathbf{j}_{nk}(\mathbf{y})}{\omega_q(\omega_{kn}-\lambda_q)}\int d^3x~\left( \mathbf{F}_q(\mathbf{x})\cdot\nabla\psi_n(\mathbf{x}) + \nabla\cdot(\mathbf{F}_q(\mathbf{x})\psi_n(\mathbf{x}))\right)
\end{equation}
A similar equation arises for the derivative with respect to an unoccupied electron orbital, except that the summation should now be over unoccupied electron orbitals.
